\chapter{General Introduction}
\label{sec:intro}

In his famous ``On the origin of species [...]'' from 1859, Charles Darwin noted
the outstanding variety of individuals within a species. Especially domesticated
speices, as he describes, show remarkable differences---imagine a bloodhound,
terrier, spaniel, and bull-dog next to one another. But also in nature such
differences occur. These difference can be passed on to offspring over
generations and accumulate to such an extent that a distinct species is formed.
It was this observation as well as a great amount of preceding research by him
and others that led Darwin to the formulation of his famous evolutionary theory.

More than 150 years later, we have understood many of the molecular mechanisms
behind this principle. We know that genetic information is encoded in DNA and
that it is subject to mutations, which are partly inheritable. These changes in
DNA create the variability of genetic material that is so essential to
evolution. Today, we are able to study the genetic differences,
which we also call genetic \emph{variants},
between human individuals or between individuals of other species and can investigate
their consequences. Many fundamental associations between the presence of
genetic variants and certain traits have been found since, for example why
red-green color blindness affects more males than females \citep{Nathans1986}.
Other traits, such as the expected height of a person, are less easily conceived
as they appear to be affected by the combination of many genetic variants
\citep{Wood2014,Marouli2017}. We also gained a much better understanding of the
causal role of genetic variation in many diseases, including Mendelian disorders
and cancer \citep{Stankiewicz2010}.

In order to study the consequences of genetic variants we require
methods to accurately detect them. Nowadays, this has been largely enabled with
the advance of DNA sequencing technologies. Based on these methods, the average
genome of many species, including humans, could be charted for the first time
\citep{Lander2001,Venter2001}, and functional units such as genes and regulatory
elements were identified in great detail \citep{Dunham2012}. Then, population
genetics studies gained further insight into the variability within the human
population \citep{Auton2015,Sudmant2015}.
At last, the identified variants could be analyzed for a potential effect on
phenotypes or disease using, for example, genome-wide association studies
\citep{Ott2015,MacArthur2017}.

Variant detection has become a standard procedure in genomics research. However,
not all types of genetic variants have been studied equally comprehensively.
Especially larger genomic rearrangments, so-called \emph{\aclp{sv}} (which are
introduced in \cref{sec:sv}), remained
difficult to ascertain using available assays, which is why their functional
role is still not as well explored as for smaller types of genetic variation.
However, \aclp{sv} are known to have extraordinary impact on our genetic
material---after all, they constitute the majority of genetic
differences within the human population \citep{Sudmant2015}.
The limitations of the current standard techniques had long been noticed and
encouraged further method development \citep{Onishi-Seebacher2011}. Over the
last couple of years, new technologies for DNA sequencing as well as new method
based on DNA sequencing have become available. These techniques hold promise to
improve the abilities to study such rearrangements. In this work, I utilize such
emerging technologies to characterize \aclp{sv} beyond what was possible previously.

In the rest of this chapter, I explain the relevant terminology around genetic
variation (\cref{sec:variation}), introduce structural variants and the biology
behind them (\cref{sec:sv}), give an overview of previous and emerging DNA
sequencing technologies (\cref{sec:sequencing}), and outline the methodology
of SV detection (\cref{sec:sv_detection}). I continue to elaborate the current
limitations of \acl{sv} detection (\cref{sec:limitations}). Then, I formulate
the goals of my research and provide an overview of the studies covered in this
dissertation (\cref{sec:motivation}).






\section{Terminology around genetic variation}
\label{sec:variation}

Alterations in the DNA of an organism, or of a cell, can arise spontaneously
via chemical or biological processes. If not repaired faithfully, they leave
traces in the genetic material that we call genetic \emph{variants}. The process
of altering the DNA seqeunce is called \emph{mutation}, but the terms mutation
and variant are often used interchangeably.
We already know the nucleotide sequence of the genomes of many species,
including humans. This sequence, which is supposed to represent the an average
of the individuals, is stored in a so-called \emph{reference} genome (also
\emph{reference assembly}, or simply \emph{reference}).
Thus, variants are usually defined as a difference to this reference. Some
variants only affect a single nucleotide, e.g. by changing a cytosine into a
thymine, and they are called \emph{\acfp{snv}}. Others variants delete or insert
nucleotides, or fully rearrange their order, as will be introduced later.

Most \explain{metazoa}{Animals. Not all animals are diploid,
    though. Bees and ants, for example, produce fully haploid males, yet still
    diploid females. Mammals, on the other hand, are believed to always be
    diploid \citep{Svartman2005}} are diploid, so they carry two non-identical
copies of each chromosome in their cells: one of maternal and one of paternal origin. Any new
variant (within a cell or organism) typically arises only on one of the two
homologous chromosomes. Such a variant is said to be \emph{heterozygous}. The
genomic locus harboring this variant exists in two different versions, which we
call \emph{alleles}. Specifically it harbors a \emph{reference allele}, which is
in agreement with the reference assembly, and
an \emph{alternative allele} that describes the non-reference variant form. A
site with exactly two alleles seen across a population is termed
\emph{bi-allelic}, but there are also sites that contain multiple different
alleles and are thus \emph{multi-allelic}. Chromosomes can contain many variants
and, depending on the detection strategies, it is often unclear which variants
reside on the same homologue. If this is known, we refer to the ensemble of
variants present along a single homologue as a haploid genotype, or short
\emph{haplotype}.

Alleles that are present in the germ line, i.e. in cells
carrying inheritable genetic material, can be propagated to offspring. This way,
an individual can end up carrying the same variant of a genomic locus on both
homologues, which makes it a \emph{homozygous} variant. Variants that are seen
more often in a population, specifically in at least 1\% of the homologoues,
are also called \emph{polymorphisms}.

When a variant is present in an individual, but not in their parents, we call it
a \emph{de novo} variant (or \emph{de novo} mutation). The mutation could
either have occurred within the
zygote during the first few cell divisions, or already beforehand in the
parental germ line. The latter can sometimes be inferred if other offspring of
those parents carries the same variant. Variants that occur not in the germ line
but in cells of the non-inheritable part of an organism are \emph{somatic}
variants. When such an affected cell undergoes repeated divisions, a somatic variant
can be present in a relevant fraction of the total cells of an individual---this
is called \emph{somatic mosaicism}. Depending on how early in development the
variant occurred (and on its effect on the fitness of the cell), it can be present in all cells of the
same lineage \citep{Youssoufian2002}. Cancer is a pathological form of somatic
mosaicism, which arises through the clonal expension of cells. \citep{Campbell2007}.


\section{Structural Variation}
\label{sec:sv}
\input{svs}


\section{DNA sequencing technologies}
\label{sec:sequencing}
\input{seq_tech}


\section{Structural variant detection}
\label{sec:sv_detection}
\input{sv_detection}


\section{Research goals and thesis overview}
\label{sec:motivation}

Studies of structural variation are still fundamentally hampered by the
shortcomings of current \sv detection methods. This especially affects studies
of balanced or complex rearrangements, which had often remained cryptic in
previous studies. In this dissertation, I aim at uncovering and further examining
\acp{sv} that had been difficult to ascertain beforehand. In order to do so, I
utilize emerging sequencing technologies and protocols---namely the
techniques introduced in \crefrange{sec:long_read_seq}{sec:strandseq}.
My work is structured into three separate research projects, in which I explore
one of the techniques, each. Below, the specific aims of each project are
outlined. A common theme throughout these projects is the design of new
bioinformatics methods for data analysis and visualization. I believe that the
full potential of novel technology can only be unlocked by the simultaneous
development of new computational approaches. I hence thrive for advancing the
state of the art of current methodology by building---and making
available---software tools that future research benefits from.

In \cref{sec:complex_invs}, I present my work for the final phase of the
1000 Genomes Project, in which my role was to validate inversion predictions in
the human population. These inversion predictions had initially remained
inconclusive in \pcr-based validation experiments. We then utilized targeted
long-read sequencing on both \pacbio and \ont MinION platforms to examine the
respective loci more closely. My immediate goal in this project was to verify
inversion calls based on long-read information. Moreover, the subsequent goal
was to further characterize these loci and investigate why previous validations
had failed. Afterwards, an optional goal was to investigate the validated
inversions deeper to understand the biological mechanisms that created
them.

In \cref{sec:balancer}, I focused on the functional aspects of structural variation.
Together with my collaborators, we set out to study the consequences of
chromosomal rearrangements on gene expression and on the three-dimensional
organization of chromosomes---the latter had gained much attention in recent
years and is introduce in depth in \cref{sec:balancer_background}. We utilized \hic
to study chromatin conformation of highly rearranged chromosomes of
\textit{Drosophila melanogaster}. A crucial initial step of the project was to
map the rearrangements and other variation present in these chromosomes. My
first goal thus was to characterize \acp{sv}, including large pericentric
inversions, which had been known only at cytogenetic resolution beforehand.
As a part of this goal, I aimed at exploring the usability of
\hic data for validation and characterization purposes. The second and more
general goal of this study was to find out whether, and how these rearrangements
affect gene expression. My specific milestone towards this goal was to robustly
determine differential gene expression between the rearranged and a
non-rearranged (wild type) chromosome. At last, this information should be
integrated with findings on chromatin structure (provided by collaborators) to
be able to conclude on the impact of rearrangements and on the mechanisms that
are involved.

In \cref{sec:mosaicatcher}, my collaborators and I explored the potential of
Strand-seq to discover \acp{sv} in single cells. While this had been previously
done for inversions, we aimed at extending this principle---for the very first
time---to at as many \sv classes as possible. Unlike previous efforts using Strand-seq,
the particular goal was to enable this detection also for variants that are present in only
a low fraction of cells.
Here, I present the current state of this ongoing endeavor. The first
goal in this study was to develop a general concept of how different \sv classes
can be revealed based on the signals available in single-cell Strand-seq data.
Then, the next goal was to prove feasibility of this concept by implementing the
approach and applying it to available as well as newly sequenced cell line data
harboring structural variation. A third important aspect was to test the
limitations of our approach in a controlled environment.

At last, \cref{sec:conclusions} concludes on the achievements resulting from my
work. This includes the specific questions of the three projects as well as a
more general view on how emerging technologies improved \sv characterization in
the three different scenarios. I then briefly review recent developments by others in
the community that occurred around the same time as my research. This should give a
broader impression on how emerging technologies are about to change studies on
structural variation and places my work within the broader context of recent
advances in the field.

